---
title: "Average Total Fertility Rate by Geographic Region (1960-2021)"
author: "Bradley Cardona"
date: "2023-06-28"
output:
  pdf_document: default
  html_document: default
  word_document: default
geometry: a4paper, margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

#### Case Study Outline:
- Title: Average Total Fertility Rate by Geographic Region (1960-2021)
- Question: What was the Average Total Fertility Rate by Geographic Region for the last 50 years?
- Dataset: [World Bank](https://data.worldbank.org/indicator/SP.DYN.TFRT.IN?end=2021&start=2021&view=map&year=2021)

## Install required packages

```{r}
library(tidyverse)  # helps wrangle data
library(lubridate)  # helps wrangle date attributes
library(ggplot2)  # helps visualize data
library(readxl)
getwd() # displays working directory

# set working directory to simplify calls to data
# setwd("~/Documents/data_analytics/fertility_rate_case_study/") 
```

## Step 1: Collect data

```{r}
fertility_rates <- read_xls("fertility_rates.xls")
head(fertility_rates)
```

## Step 2: Wrangle data, look for any incongruencies 

```{r}
colnames(fertility_rates)  # Column names
nrow(fertility_rates)  # Number of rows
dim(fertility_rates)  # Dimensionss
head(fertility_rates)  # First 6 rows of data frame
str(fertility_rates)  # Columns and respective data types (numeric, character, etc)
summary(fertility_rates)  # Statistical summary of data. Mainly for numerics

# delete two columns: "Indicator Name" and "Indicator Code"
fertility_rates <- fertility_rates %>% 
  select(-c("Indicator Name", "Indicator Code"))
```

## Step 3: Clean up data and prepare for analysis 

Inspect the new table that has been created

```{r}
colnames(fertility_rates)  # Column names
nrow(fertility_rates)  # Number of rows
dim(fertility_rates)  # Dimensions
head(fertility_rates)  # First 6 rows of data frame
str(fertility_rates)  # Columns and respective data types (numeric, character, etc)
summary(fertility_rates)  # Statistical summary of data. Mainly for numerics
```

## Step 4: Conduct descriptive analysis

```{r}
summary(fertility_rates$"1960")

# Compare Average Total Fertility rates in 1960 vs. Average Total Fertility rates in 2021
aggregate(fertility_rates$"1960" ~ fertility_rates$"2021", FUN = mean)
# Compare median fertility rates in 1960 vs. median fertility in 2021
aggregate(fertility_rates$"1960" ~ fertility_rates$"2021", FUN = median)
# Compare maximum fertility rates in 1960 vs. maximum fertility rates in 2021
aggregate(fertility_rates$"1960" ~ fertility_rates$"2021", FUN = max)
# Compare minimum fertility rates in 1960 vs. minimum fertility rates in 2021
aggregate(fertility_rates$"1960" ~ fertility_rates$"2021", FUN = min)

# Find Average Total Fertility rate by Country Name in 1960 vs. 2021. 
# Assign the result to avg_fertility_rate.
avg_fertility_rate <- aggregate(fertility_rates$"1960" 
                                ~ fertility_rates$"2021" 
                                + fertility_rates$"Country Name", 
                                FUN = mean)
```


## Step 5: Create figures

### Figure 1
```{r fig.align='center'}
# Create a scatter plot of the data to find general pattern of fertility rates
ggplot(avg_fertility_rate, aes(x = `fertility_rates$"2021"`, y = `fertility_rates$"1960"`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Average Total Fertility Rate in 2021") +
  ylab("Average Total Fertility Rate in 1960") +
  ggtitle("Average Total Fertility Rate by Country (1960 vs. 2021)")

```

### Figure 2
```{r fig.align='center'}
# Find unique country names in fertility_rates table
country_names <- unique(fertility_rates$`Country Name`)
sorted_country_names <- sort(country_names)
sorted_country_names
```

```{r}
# Filter the dataset to include only European countries
european_countries <- c("Albania", "Andorra", "Austria", "Belarus", "Belgium", 
                        "Bosnia and Herzegovina", "Bulgaria", "Channel Islands", 
                        "Croatia", "Cyprus", "Czechia", "Denmark", "Estonia", 
                        "Faroe Islands", "Finland", "France", "Germany", 
                        "Gibraltar", "Greece", "Hungary", "Iceland", "Ireland",
                        "Isle of Man", "Italy", "Latvia", "Liechtenstein",
                        "Lithuania", "Luxembourg", "Malta", "Moldova",
                        "Monaco", "Montenegro", "Netherlands",
                        "North Macedonia", "Norway", "Poland",
                        "Portugal", "Romania", "Russian Federation",
                        "San Marino", "Serbia","Slovakia","Slovenia","Spain",
                        "Sweden","Switzerland","Ukraine","United Kingdom",
                        "Vatican City")


# Replace with the list of European countries
fertility_rates_europe <- fertility_rates %>%
  filter(`Country Name` %in% european_countries)

# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_europe %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Create the plot with custom x-axis labels
p <- ggplot(fertility_rates_long, 
            aes(x = Year, 
                y = Fertility_Rate, 
                color = `Country Name`, 
                group = `Country Name`)) +
  geom_line() +
  xlab("Year") +
  ylab("Fertility Rate") +
  ggtitle("Average Total Fertility Rate in European Countries (1960-2021)") +
  scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  guides(color = guide_legend(ncol = 3))

p 

# Save the plot to a file with a resolution of 400 dpi
ggsave("Documents/data_analytics/fertility_rate_case_study/figures/fertility_rates_europe.png", 
       plot = p, 
       width = 10, 
       dpi = 1000)

```


### Figure 3
```{r fig.align='center', fig.height = 8}
# Filter the dataset to include only African countries
african_countries <- c("Angola", "Benin", "Burkina Faso", "Burundi", "Cabo Verde",
                       "Cameroon", "Central African Republic", "Chad", "Comoros",
                       "Congo, Dem. Rep.", "Congo, Rep.", "Cote d'Ivoire",
                       "Djibouti", "Egypt, Arab Rep.", "Equatorial Guinea",
                       "Eritrea", "Eswatini", "Ethiopia", "Gabon", "Gambia, The",
                       "Ghana", "Guinea", "Guinea-Bissau", "Kenya", "Lesotho",
                       "Liberia", "Libya", "Madagascar", "Malawi", "Mali",
                       "Mauritania", "Mauritius", "Morocco", "Mozambique",
                       "Namibia", "Niger", "Nigeria", "Rwanda",
                       "Sao Tome and Principe", "Senegal", 
                       "Seychelles","Sierra Leone","Somalia","South Africa",
                       "South Sudan","Sudan","Tanzania","Togo","Tunisia",
                       "Uganda","Zambia","Zimbabwe")


# Replace with the list of African countries
fertility_rates_africa <- fertility_rates %>%
  filter(`Country Name` %in% african_countries)

# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_africa %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Create the plot with custom x-axis labels
p <- ggplot(fertility_rates_long, 
            aes(x = Year, 
                y = Fertility_Rate, 
                color = `Country Name`, 
                group = `Country Name`)) +
  geom_line() +
  xlab("Year") +
  ylab("Fertility Rate") +
  ggtitle("Average Total Fertility Rate in African Countries (1960-2021)") +
  scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  guides(color = guide_legend(ncol = 3))

p 

# Save the plot to a file with a resolution of 400 dpi
ggsave("Documents/data_analytics/fertility_rate_case_study/figures/fertility_rates_africa.png",
       plot = p,
       width = 20,
       dpi = 1000)

```


### Figure 4
```{r fig.align='center', fig.height = 8}
# Filter the dataset to include only Asian countries
asian_countries <- c("Afghanistan", "Armenia", "Azerbaijan", "Bahrain", "Bangladesh",
                     "Bhutan", "Brunei Darussalam", "Cambodia", "China", "Georgia",
                     "Hong Kong SAR, China", "India", "Indonesia", "Iran, Islamic Rep.",
                     "Iraq", "Israel", "Japan", "Jordan", "Kazakhstan",
                     "Korea, Dem. People's Rep.", "Korea, Rep.", "Kuwait",
                     "Kyrgyz Republic", "Lao PDR", "Lebanon",
                     "Macao SAR, China", "Malaysia", "Maldives",
                     "Mongolia", "Myanmar", "Nepal",
                     "Oman", "Pakistan","Palestine","Philippines","Qatar",
                     "Russian Federation","Saudi Arabia","Singapore","Sri Lanka",
                     "Syrian Arab Republic","Taiwan, China","Tajikistan","Thailand",
                     "Timor-Leste","Turkey","Turkmenistan","United Arab Emirates",
                     "Uzbekistan","Vietnam","Yemen, Rep.")


# Replace with the list of Asian countries
fertility_rates_asia <- fertility_rates %>%
  filter(`Country Name` %in% asian_countries)

# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_asia %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Create the plot with custom x-axis labels
p <- ggplot(fertility_rates_long, 
            aes(x = Year, 
                y = Fertility_Rate, 
                color = `Country Name`, 
                group = `Country Name`)) +
  geom_line() +
  xlab("Year") +
  ylab("Fertility Rate") +
  ggtitle("Average Total Fertility Rate in Asian Countries (1960-2021)") +
  scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  guides(color = guide_legend(ncol = 3))

p 

# Save the plot to a file with a resolution of 400 dpi
ggsave("Documents/data_analytics/fertility_rate_case_study/figures/fertility_rates_asia.png",
       plot = p,
       width = 10,
       dpi = 1000)

```


### Figure 5
```{r fig.align='center', fig.height = 8}

# Filter the dataset to include only North American countries
north_american_countries <- c("Antigua and Barbuda", "Bahamas, The", "Barbados",
                              "Belize", "Canada", "Costa Rica", "Cuba", "Dominica",
                              "Dominican Republic", "El Salvador", "Greenland",
                              "Grenada", "Guatemala", "Haiti", "Honduras",
                              "Jamaica", "Mexico", "Nicaragua", "Panama",
                              "St. Kitts and Nevis", "St. Lucia",
                              "St. Vincent and the Grenadines",
                              "Trinidad and Tobago", "United States")


# Replace with the list of North American countries
fertility_rates_north_america <- fertility_rates %>%
  filter(`Country Name` %in% north_american_countries)

# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_north_america %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Create the plot with custom x-axis labels
p <- ggplot(fertility_rates_long, 
            aes(x = Year, 
                y = Fertility_Rate, 
                color = `Country Name`, 
                group = `Country Name`)) +
  geom_line() +
  xlab("Year") +
  ylab("Fertility Rate") +
  ggtitle("Average Total Fertility Rate in North American Countries (1960-2021)") +
  scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  guides(color = guide_legend(ncol = 3))

p 

# Save the plot to a file with a resolution of 400 dpi
ggsave("Documents/data_analytics/fertility_rate_case_study/figures/fertility_rates_north_america.png",
       plot = p,
       width = 10,
       dpi = 1000)

```


### Figure 6
```{r fig.align='center', fig.height = 8}

# Filter the dataset to include only South American countries
south_american_countries <- c("Argentina", "Bolivia", "Brazil", "Chile",
                              "Colombia", "Ecuador", "Guyana", "Paraguay",
                              "Peru", "Suriname", "Uruguay", "Venezuela")


# Replace with the list of South American countries
fertility_rates_south_america <- fertility_rates %>%
  filter(`Country Name` %in% south_american_countries)

# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_south_america %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Create the plot with custom x-axis labels
p <- ggplot(fertility_rates_long, 
            aes(x = Year, 
                y = Fertility_Rate, 
                color = `Country Name`, 
                group = `Country Name`)) +
  geom_line() +
  xlab("Year") +
  ylab("Fertility Rate") +
  ggtitle("Average Total Fertility Rate in South American Countries (1960-2021)") +
  scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  guides(color = guide_legend(ncol = 3))

p 

# Save the plot to a file with a resolution of 400 dpi
ggsave("Documents/data_analytics/fertility_rate_case_study/figures/fertility_rates_south_america.png",
       plot = p,
       width = 10,
       dpi = 1000)


```


### Figure 7
```{r fig.align='center', fig.height = 8}

# Filter the dataset to include only Oceanic countries
oceanic_countries <- c("Australia", "Fiji", "Kiribati", "Marshall Islands",
                       "Micronesia, Fed. Sts.", "Nauru", "New Zealand", "Palau",
                       "Papua New Guinea", "Samoa", "Solomon Islands", "Tonga",
                       "Tuvalu", "Vanuatu")


# Replace with the list of Oceanic countries
fertility_rates_oceania <- fertility_rates %>%
  filter(`Country Name` %in% oceanic_countries)

# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_oceania %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Create the plot with custom x-axis labels
p <- ggplot(fertility_rates_long, 
            aes(x = Year, 
                y = Fertility_Rate, 
                color = `Country Name`, 
                group = `Country Name`)) +
  geom_line() +
  xlab("Year") +
  ylab("Fertility Rate") +
  ggtitle("Average Total Fertility Rate in Oceanic Countries (1960-2021)") +
  scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom") +
  guides(color = guide_legend(ncol = 3))

p 

# Save the plot to a file with a resolution of 400 dpi
ggsave("Documents/data_analytics/fertility_rate_case_study/figures/fertility_rates_oceania.png",
       plot = p,
       width = 10,
       dpi = 1000)

```


#### Average Total Fertility Rate by Geographic Region (1960-2021)
```{r fig.align='center', fig.height = 8}

# Combine the data for all regions into one data frame
fertility_rates_all_regions <- bind_rows(
  fertility_rates_africa %>% mutate(Region = "Africa"),
  fertility_rates_asia %>% mutate(Region = "Asia"),
  fertility_rates_europe %>% mutate(Region = "Europe"),
  fertility_rates_north_america %>% mutate(Region = "North America"),
  fertility_rates_oceania %>% mutate(Region = "Oceania"),
  fertility_rates_south_america %>% mutate(Region = "South America")
)

# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_all_regions %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Create the plot with custom x-axis labels and facets
p1 <- ggplot(fertility_rates_long, 
             aes(x = Year, 
                 y = Fertility_Rate, 
                 color = `Country Name`, 
                 group = `Country Name`)) +
  geom_line() +
  xlab("Year") +
  ylab("Fertility Rate") +
  ggtitle("Average Total Fertility Rate by Region (1960-2021)") +
  scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  facet_wrap(~ Region)

p1

# Save the plot with the graphs to a file with a resolution of 400 dpi
ggsave("Documents/data_analytics/fertility_rate_case_study/figures/fertility_rates_all_regions_graphs.png",
       plot = p1,
       width = 10,
       dpi = 1000)

# Create a separate plot for the legend
```

## Step 6: Export data, create Tableau dashboard

```{r }
# Overall, it seems that many countries are experiencing a pronounced downward 
# trend in terms of average total fertility rate. To make this more clear, 
# let's export fertility_rates_long.csv to Tableau to create an interactive 
# dashboard. 

write.csv(fertility_rates_long, 
          "Documents/data_analytics/fertility_rate_case_study/fertility_rates_all_regions.csv",
          row.names = FALSE)

```

## Step 7: Explore Tableau dashboard
- Click [here](https://public.tableau.com/app/profile/bradley.cardona/viz/AverageFertilityRatebyRegion1960-2021/Dashboard2) to see the corresponding Tableau dashboard.



