---
title: "Overpopulation vs. Underpopulation: Insights from Total Fertility Rate Data (1960-2021)"
author: "Bradley Cardona"
date: "2023-06-28"
output:
  pdf_document: default
  html_document: default
  word_document: default
geometry: a4paper, margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, fig.width = 14, fig.height = 10)

```

#### Case Study Outline:
- Motivating Question: Which issue will be more significant in the coming decades, overpopulation or underpopulation?
- Hypothesis: In the coming decades, most countries will need to address an *underpopulation* problem -- not an overpopulation one.
- Data set: For this case study, I used a [data set](https://data.worldbank.org/indicator/SP.DYN.TFRT.IN?end=2021&start=2021&view=map&year=2021) from The World Bank. It contains a list of countries, in addition to their respective total fertility rates from 1960 to 2021. 


#### Abstract: 
- The Total Fertility Rate (TFR) is a measure of the average number of children that would be born to a woman over her lifetime if she were to experience the exact current age-specific fertility rates through her lifetime and were to live from birth until the end of her reproductive life. It is often said that the world is experiencing, or will experience an overpopulation problem. Due to the dramatic drop in TFRs across nearly every country for the last several decades, we hypothesize that most countries will need to address an underpopulation problem -- not an overpopulation one.

## Install required packages

```{r}
library(tidyverse)  # helps wrangle data
library(lubridate)  # helps wrangle date attributes
library(ggplot2)  # helps visualize data
library(readxl) # helps read data from Excel files 
library(scales) # helps scale data, format labels, and create color palettes
library(dplyr) # helps manipulate data frames

getwd() # displays working directory

# set working directory to simplify calls to data
# setwd("~/Documents/data_analytics/fertility_rate_case_study/") 
```

## Step 1: Collect data

```{r}
fertility_rates <- read_xls("fertility_rates.xls")
head(fertility_rates)
```

## Step 2: Wrangle data, look for any incongruencies 

```{r}
colnames(fertility_rates)  # Column names
nrow(fertility_rates)  # Number of rows
dim(fertility_rates)  # Dimensionss
head(fertility_rates)  # First 6 rows of data frame
str(fertility_rates)  # Columns and respective data types (numeric, character, etc)
summary(fertility_rates)  # Statistical summary of data. Mainly for numerics

# delete two columns: "Indicator Name" and "Indicator Code"
fertility_rates <- fertility_rates %>% 
  select(-c("Indicator Name", "Indicator Code"))
```

## Step 3: Clean up data and prepare for analysis 

Inspect the new table that has been created

```{r}
colnames(fertility_rates)  # Column names
nrow(fertility_rates)  # Number of rows
dim(fertility_rates)  # Dimensions
head(fertility_rates)  # First 6 rows of data frame
str(fertility_rates)  # Columns and respective data types (numeric, character, etc)
summary(fertility_rates)  # Statistical summary of data. Mainly for numerics
```

## Step 4: Conduct descriptive analysis

```{r}

# Compare total fertility rates in 1960 vs. total fertility rates in 2021
aggregate(fertility_rates$"1960" ~ fertility_rates$"2021", FUN = mean)
# Compare median fertility rates in 1960 vs. median fertility in 2021
aggregate(fertility_rates$"1960" ~ fertility_rates$"2021", FUN = median)
# Compare maximum fertility rates in 1960 vs. maximum fertility rates in 2021
aggregate(fertility_rates$"1960" ~ fertility_rates$"2021", FUN = max)
# Compare minimum fertility rates in 1960 vs. minimum fertility rates in 2021
aggregate(fertility_rates$"1960" ~ fertility_rates$"2021", FUN = min)

# Find Total Fertility rate by Country Name in 1960 vs. 2021. 
# Assign the result to avg_fertility_rate.
avg_fertility_rate <- aggregate(fertility_rates$"1960" 
                                ~ fertility_rates$"2021" 
                                + fertility_rates$"Country Name", 
                                FUN = mean)
```


## Step 5: Create figures

```{r fig.align='center'}
# Figure 1: Total Fertility Rate by Country (1960 vs. 2021)
ggplot(avg_fertility_rate, aes(x = `fertility_rates$"2021"`, y = `fertility_rates$"1960"`)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Total Fertility Rate in 2021") +
  ylab("Total Fertility Rate in 1960") +
  ggtitle("Figure 1. Total Fertility Rate by Country (1960 vs. 2021)")
```
Each data point in Figure 1 represents a different country. These data points suggest that there has been a considerable decrease in total fertility rates from 1960 to 2021. Let's now take a look at this trend on a region-by-region basis.

```{r fig.align='center'}
# Find unique country names in fertility_rates table
country_names <- unique(fertility_rates$`Country Name`)
sorted_country_names <- sort(country_names)
sorted_country_names
```


```{r}

# Define the lists of countries by region
european_countries <- c("Albania", "Andorra", "Austria", "Belarus", "Belgium", 
                        "Bosnia and Herzegovina", "Bulgaria", "Channel Islands", 
                        "Croatia", "Cyprus", "Czechia", "Denmark", "Estonia", 
                        "Faroe Islands", "Finland", "France", "Germany", 
                        "Gibraltar", "Greece", "Hungary", "Iceland", "Ireland",
                        "Isle of Man", "Italy", "Latvia", "Liechtenstein",
                        "Lithuania", "Luxembourg", "Malta", "Moldova",
                        "Monaco", "Montenegro", "Netherlands",
                        "North Macedonia", "Norway", "Poland",
                        "Portugal", "Romania", "Russian Federation",
                        "San Marino", "Serbia","Slovakia","Slovenia","Spain",
                        "Sweden","Switzerland","Ukraine","United Kingdom",
                        "Vatican City")

african_countries <- c("Angola", "Benin", "Burkina Faso", "Burundi", "Cabo Verde",
                       "Cameroon", "Central African Republic", "Chad", "Comoros",
                       "Congo, Dem. Rep.", "Congo, Rep.", "Cote d'Ivoire",
                       "Djibouti", "Egypt, Arab Rep.", "Equatorial Guinea",
                       "Eritrea", "Eswatini", "Ethiopia", "Gabon", "Gambia, The",
                       "Ghana", "Guinea", "Guinea-Bissau", "Kenya", "Lesotho",
                       "Liberia", "Libya", "Madagascar", "Malawi", "Mali",
                       "Mauritania", "Mauritius", "Morocco", "Mozambique",
                       "Namibia", "Niger", "Nigeria", "Rwanda",
                       "Sao Tome and Principe", "Senegal", 
                       "Seychelles","Sierra Leone","Somalia","South Africa",
                       "South Sudan","Sudan","Tanzania","Togo","Tunisia",
                       "Uganda","Zambia","Zimbabwe")

asian_countries <- c("Afghanistan", "Armenia", "Azerbaijan", "Bahrain", "Bangladesh",
                     "Bhutan", "Brunei Darussalam", "Cambodia", "China", "Georgia",
                     "Hong Kong SAR, China", "India", "Indonesia", "Iran, Islamic Rep.",
                     "Iraq", "Israel", "Japan", "Jordan", "Kazakhstan",
                     "Korea, Dem. People's Rep.", "Korea, Rep.", "Kuwait",
                     "Kyrgyz Republic", "Lao PDR", "Lebanon",
                     "Macao SAR, China", "Malaysia", "Maldives",
                     "Mongolia", "Myanmar", "Nepal",
                     "Oman", "Pakistan","Palestine","Philippines","Qatar",
                     "Russian Federation","Saudi Arabia","Singapore","Sri Lanka",
                     "Syrian Arab Republic","Taiwan, China","Tajikistan","Thailand",
                     "Timor-Leste","Turkey","Turkmenistan","United Arab Emirates",
                     "Uzbekistan","Vietnam","Yemen, Rep.")

north_american_countries <- c("Antigua and Barbuda", "Bahamas, The", "Barbados",
                              "Belize", "Canada", "Costa Rica", "Cuba", "Dominica",
                              "Dominican Republic", "El Salvador", "Greenland",
                              "Grenada", "Guatemala", "Haiti", "Honduras",
                              "Jamaica", "Mexico", "Nicaragua", "Panama",
                              "St. Kitts and Nevis", "St. Lucia",
                              "St. Vincent and the Grenadines",
                              "Trinidad and Tobago", "United States")

south_american_countries <- c("Argentina", "Bolivia", "Brazil", "Chile",
                              "Colombia", "Ecuador", "Guyana", "Paraguay",
                              "Peru", "Suriname", "Uruguay", "Venezuela")

oceanic_countries <- c("Australia", "Fiji", "Kiribati", "Marshall Islands",
                       "Micronesia, Fed. Sts.", "Nauru", "New Zealand", "Palau",
                       "Papua New Guinea", "Samoa", "Solomon Islands", "Tonga",
                       "Tuvalu", "Vanuatu")

# Define a function to generate a plot for a given region
generate_plot <- function(region, countries, data, country_col, file_path) {
  # Filter the dataset to include only countries in the given region
  fertility_rates_region <- data %>%
    filter(!!sym(country_col) %in% countries)
  
  # Reshape the data from wide to long format
  fertility_rates_long <- fertility_rates_region %>%
    pivot_longer(cols = `1960`:`2021`,
                 names_to = "Year",
                 values_to = "Fertility_Rate")
  
  # Create the plot with custom x-axis labels
  p <- ggplot(fertility_rates_long, 
              aes(x = Year, 
                  y = Fertility_Rate, 
                  color = !!sym(country_col), 
                  group = !!sym(country_col))) +
    geom_line() +
    xlab("Year") +
    ylab("Total Fertility Rate") +
    ggtitle(paste0("Total Fertility Rate in ", region, 
                   " (1960-2021)")) +
    scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom") +
    guides(color = guide_legend(ncol = 3))
  
  # Save the plot to a file with a resolution of 400 dpi
  ggsave(paste0(file_path, "/fertility_rates_", 
                tolower(region), ".png"), 
         plot = p, 
         width = 14,
         height = 10,
         dpi = 200)
  
  return(p)
}

# Generate a plot for each region

### Figure 2 (European Countries)
generate_plot("Europe", european_countries, fertility_rates, "Country Name",
              "~/Documents/data_analytics/fertility_rate_case_study/figures")

### Figure 3 (African Countries)
generate_plot("Africa", african_countries, fertility_rates, "Country Name",
              "~/Documents/data_analytics/fertility_rate_case_study/figures")

### Figure 4 (Asian Countries)
generate_plot("Asia" , asian_countries, fertility_rates, "Country Name",
              "~/Documents/data_analytics/fertility_rate_case_study/figures")

### Figure 5 (North American Countries)
generate_plot("North America" , north_american_countries, fertility_rates, "Country Name",
              "~/Documents/data_analytics/fertility_rate_case_study/figures")

### Figure 6 (South American Countries)
generate_plot("South America" , south_american_countries, fertility_rates, "Country Name",
              "~/Documents/data_analytics/fertility_rate_case_study/figures")

### Figure 7 (Oceanic Countries)
generate_plot("Oceania" , oceanic_countries, fertility_rates, "Country Name",
              "~/Documents/data_analytics/fertility_rate_case_study/figures")
```

Now that we have an idea of the general trend for each region, let's aggregate these figures into one facet plot.

```{r fig.align='center', fig.height = 8}

# Replace with the list of African countries
fertility_rates_africa <- fertility_rates %>%
  filter(`Country Name` %in% african_countries)

# Replace with the list of Asian countries
fertility_rates_asia <- fertility_rates %>%
  filter(`Country Name` %in% asian_countries)

# Replace with the list of European countries
fertility_rates_europe <- fertility_rates %>%
  filter(`Country Name` %in% european_countries)

# Replace with the list of North American countries
fertility_rates_north_america <- fertility_rates %>%
  filter(`Country Name` %in% north_american_countries)

# Replace with the list of South American countries
fertility_rates_south_america <- fertility_rates %>%
  filter(`Country Name` %in% south_american_countries)

# Replace with the list of Oceanic countries
fertility_rates_oceania <- fertility_rates %>%
  filter(`Country Name` %in% oceanic_countries)



# Combine the data for all regions into one data frame
fertility_rates_all_regions <- bind_rows(
  fertility_rates_africa %>% mutate(Region = "Africa"),
  fertility_rates_asia %>% mutate(Region = "Asia"),
  fertility_rates_europe %>% mutate(Region = "Europe"),
  fertility_rates_north_america %>% mutate(Region = "North America"),
  fertility_rates_oceania %>% mutate(Region = "Oceania"),
  fertility_rates_south_america %>% mutate(Region = "South America")
)

# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_all_regions %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Create the plot with custom x-axis labels and facets
p1 <- ggplot(fertility_rates_long, 
             aes(x = Year, 
                 y = Fertility_Rate, 
                 color = `Country Name`, 
                 group = `Country Name`)) +
  geom_line() +
  xlab("Year") +
  ylab("Fertility Rate") +
  ggtitle("Total Fertility Rate by Geographic Region (1960-2021)") +
  scale_x_discrete(breaks = seq(1960, 2021, by = 5)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  facet_wrap(~ Region)

p1

# Save the plot with the graphs to a file with a resolution of 400 dpi
ggsave("~/Documents/data_analytics/fertility_rate_case_study/figures/fertility_rates_all_regions_graphs.png",
       plot = p1,
       width = 10,
       dpi = 1000)
```
This facet plot shows clearly a downward trend of total fertility rates in the last several decades; to make this data more interactive, let's create a Tableau dashboard.

## Step 6: Export data, create Tableau dashboard

```{r }
write.csv(fertility_rates_long, 
          "~/Documents/data_analytics/fertility_rate_case_study/fertility_rates_all_regions.csv",
          row.names = FALSE)
```

## Step 7: Explore Tableau dashboard
- Click [here](https://public.tableau.com/app/profile/bradley.cardona/viz/AverageFertilityRatebyRegion1960-2021/Dashboard2) to interactive with this case study's dashboard.

## Step 8: Identify pattern
The Tableau dashboard shows that, for nearly every country, there has been a sharp decrease in total fertility rates from 1960 to 2021. Let's quantify this result even further by finding the average percentage change in total fertility rate by country in each region from 1960 to 2021. If the average percentage change is positive for a given country, we will label it with a green circle; otherwise, we will label it with a red circle. 


```{r}
# Reshape the data from wide to long format
fertility_rates_long <- fertility_rates_all_regions %>%
  pivot_longer(cols = `1960`:`2021`,
               names_to = "Year",
               values_to = "Fertility_Rate")

# Replace spaces with underscores in column names
colnames(fertility_rates_long) <- gsub(" ", "_", colnames(fertility_rates_long))

# Add a pct_change column to the data frame
fertility_data <- fertility_rates_long %>%
  group_by(Country_Name) %>%
  mutate(pct_change = (Fertility_Rate - lag(Fertility_Rate)) / lag(Fertility_Rate) * 100) %>%
  ungroup()

plots <- fertility_data %>%
  group_by(Region, Country_Name) %>%
  summarise(mean_pct_change = mean(pct_change, na.rm = TRUE)) %>%
  ungroup() %>%
  split(.$Region) %>%
  map(~ .x[order(.x$mean_pct_change),] %>%
    ggplot(aes(x = reorder(Country_Name, mean_pct_change), y = mean_pct_change)) +
    geom_point(shape = 21, size = 3, aes(fill = ifelse(mean_pct_change > 0, "Above 0", "Below 0"))) +
    scale_fill_manual(values = c("Above 0" = "green", "Below 0" = "red")) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
    labs(x = "Country Name", y = "Average Percentage Change in Total Fertility Rate", title = paste("Average Percentage Change in Total Fertility Rate by Country in", unique(.x$Region), " (1960-2021)")))

# Print the plots
plots
```

#### Conclusion:
These figures show that that the average percentage change in total fertility rates is negative for nearly every country from 1960 to 2021. This supports our initial hypothesis that, in the coming decades, many countries will need to address an underpopulation problem -- not an overpopulation one.   
